---
title: "Survival Models and The Leukemia Dataset"
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 80
---

# Dataset

In this document, we illustrate the use of the proportional hazards model
applied with BART to survival analysis. The dataset we are interested is
described in the `spBayesSurv` package as follows:

> A dataset on the survival of acute myeloid leukemia in 1,043 patients, first
> analyzed by Henderson et al. (2002). It is of interest to investigate possible
> spatial variation in survival after accounting for known subject-specific
> prognostic factors, which include age, sex, white blood cell count (wbc) at
> diagnosis, and the Townsend score (tpi) for which higher values indicates less
> affluent areas. Both exact residential locations of all patients and their
> administrative districts (24 districts that make up the whole region) are
> available.

We model this data using both the proportional hazards (PH) and non-proportional
hazards (NPH) versions of the survival model described in the paper. The PH
variant of the model is given by

$$
  \lambda(t \mid x) = \lambda_0(t) \, e^{r(x)}
$$

where $\lambda(t \mid x)$ is the *hazard function*
$\frac{d}{dt} - \log S(t \mid x) = \frac{d}{dt} - \log \Pr(T_i > t \mid X_i = x)$
and $\lambda_0(t)$ is modeled using a piecewise-constant hazard function.

First we load the required packages:

```{r loadpacks, message=FALSE}
library(loo)
library(ggdist)
library(spBayesSurv)
library(mgcv)
library(rpart)
library(rpart.plot)
library(possum)
library(BART4RS)
library(tidybayes)
library(tidyverse)
theme_set(theme_bw())
SEED <- digest::digest2int("analyze leukemia")
```

```{r loaddata}
data("LeukSurv")
head(LeukSurv)
```

We use a slightly modified version of this dataset, with the covariates of
interest scaled:

```{r otherleuk}
data("leuk_data")
head(leuk_data)
```

We first set a seed so that our results can be replicated, then we use the
`coxpe_bart` function in the `BART4RS` package to fit the model:

```{r fitmodel, cache=TRUE}
set.seed(SEED + 1)

fitted_coxpe <-
  BART4RS::coxpe_bart(
    formula = Surv(event_time, status) ~ age + sex + wbc + tpi,
    data = leuk_data,
    num_burn = 2500,
    num_save = 2500,
    num_thin = 10
  )
```

# Posterior Summarization

Next, we use the `possum` package (with some additional, optional, code for
improving the visualization) to compute the posterior projection of the fitted
BART model for $r(x)$ onto an additive model. We also compute projections
leaving one predictor out at a time to help assess which predictors are most
relevant in forming prediction.

```{r possumfits, cache = TRUE}

r <- fitted_coxpe$r_train
r_hat <- fitted_coxpe$r_train_hat

possum_summary <-
  possum::additive_summary(
    r_hat ~ s(age) + factor(sex) + s(logwbc) + s(tpi),
    fhatSamples = t(r),
    fhat = r_hat,
    df = LeukSurv %>% mutate(logwbc = log(1 + wbc))
  )

possum_summary_nosex <-
  possum::additive_summary(
    r_hat ~ s(age) + s(wbc) + s(tpi),
    fhatSamples = t(r),
    fhat = r_hat,
    df = LeukSurv
  )

possum_summary_noage <-
  possum::additive_summary(
    r_hat ~ s(wbc) + factor(sex) + s(tpi),
    fhatSamples = t(r),
    fhat = r_hat,
    df = LeukSurv
  )

possum_summary_notpi <-
  possum::additive_summary(
    r_hat ~ s(wbc) + factor(sex) + s(age),
    fhatSamples = t(r),
    fhat = r_hat,
    df = LeukSurv
  )

possum_summary_nowbc <-
  possum::additive_summary(
    r_hat ~ s(tpi) + factor(sex) + s(age),
    fhatSamples = t(r),
    fhat = r_hat,
    df = LeukSurv
  )
```

The following function is a modification of the `additive_summary_plot` function
in `possum` that makes slightly prettier pictures.

```{r additivesummary}
additive_summary_plot_2 <-  function (additive_summary,
                                      ribbonFill = "grey80",
                                      windsor = NA)
{
  temp <- additive_summary$gamDf %>%
    mutate(term = case_when(
      term == "age" ~ "Age",
      term == "tpi" ~ "TPI",
      term == "wbc" ~ "WBC",
      TRUE           ~ term
    ))
  if (!is.na(windsor)) {
    if (!("quant" %in% colnames(temp))) {
      stop("Quantiles not supplied")
    }
    temp <- temp %>% filter(quant > windsor/2 & quant < 1 -
                              windsor/2)
    glimpse(temp)
  }
  temp %>% distinct() %>%
    ggplot() + geom_hline(yintercept = 0) +
    geom_ribbon(aes(x_j, ymin = fx_j_lo, ymax = fx_j_hi),
                fill = ribbonFill, alpha = 0.5) +
    geom_line(aes(x_j, fx_j_mean), col = "firebrick3") +
    geom_rug(aes(x_j, fx_j_mean), sides = "b", alpha = 0.25) +
    facet_wrap(~term, scale = "free") +
    labs(x = ("term"), y = ("Partial effect"))
}
```

We note print the summary. We do not visualize the posterior effect of `sex` in
this case because it turns out not to be very relevant, as we will see when we
look at the summary $R^2$'s for the models:

```{r add_sum, fig.align='center', dev='tikz', fig.height=3}
add_sum <- additive_summary_plot_2(possum_summary) + xlab("") + theme_bw()
plot(add_sum)
```

Each of the three predictors examined here (age, white blood cell count, and
Townsend index) have an increasing relationship with $r(x)$.

Next, we compare the summary $R^2$ for each of the summary models; recall from
@woody2020 that this is given by the quantity

$$
  R^2 = \frac{\sum_i \{r(X_i) - \widetilde r(X_i)\}}{\sum_I \{r(X_i) - \bar r\}^2}
$$

where $\widetilde r = \arg \min_{q \in \mathcal Q} \|r - q\|$ and
$\bar r = \frac{1}{N} \sum_i r(X_i)$. This quantity represents the proportion of
the variability in $r(x)$ that can be explained by the projection onto the
simpler summary model $\mathcal Q$. Doing this, we obtain the following
posterior distributions for the summary $R^2$:

```{r rsqplt, fig.align='center', dev='tikz'}
rsq_df <- data.frame(Rsq = c(possum_summary$summaryRsq,
                             possum_summary_nosex$summaryRsq,
                             possum_summary_noage$summaryRsq,
                             possum_summary_notpi$summaryRsq,
                             possum_summary_nowbc$summaryRsq),
                     Model = rep(c("All",
                                   "Without Sex",
                                   "Without Age",
                                   "Without TPI",
                                   "Without WBC"),
                                 each = 2500))

p_rsq <- ggplot(rsq_df) +
  geom_density(aes(x = Rsq, y = after_stat(density), fill = Model),
                 color = 'white', alpha = 0.3) +
  xlab("Summary $R^2$") +
  ylab("Density") +
  theme_bw()

plot(p_rsq)

```

Lastly, we'll plot the baseline hazard function over time; we see that it is
decreasing over time, with individuals having a very high risk near baseline and
much lower risk if they survive sufficiently long.

```{r hazplot, dev="tikz", fig.height=3, fig.width=7}
g <- c(lag(fitted_coxpe$time_grid)[-c(1:2)], 20)
adjusted_hazard <- fitted_coxpe$base_haz

haz_df <- data.frame(haz_hat = colMeans(adjusted_hazard),
                     steps = fitted_coxpe$time_grid[-1],
                     widths = g,
                     lower = apply(adjusted_hazard, 2, \(x) quantile(x, 0.025)),
                     upper = apply(adjusted_hazard, 2, \(x) quantile(x, 00.975)))

p_1 <- ggplot(haz_df, aes(x = steps, y = haz_hat)) + geom_step() +
  geom_step(aes(y = lower), lty = 2) + geom_step(aes(y = upper), lty = 2) +
  xlab("Time") + ylab("") + xlim(1, 14) + ylim(0, 1.15) + theme_bw()
p_2 <- ggplot(haz_df, aes(x = steps, y = haz_hat)) + geom_step() +
  geom_step(aes(y = lower), lty = 2) + geom_step(aes(y = upper), lty = 2) +
  xlab("Time") + ylab("") + xlim(0, 1) + theme_bw()
p_3 <- ggplot(haz_df, aes(x = steps, y = haz_hat)) + geom_step() +
  geom_step(aes(y = lower), lty = 2) + geom_step(aes(y = upper), lty = 2) +
  xlab("Time") + ylab("Hazard") + theme_bw()

hazplot <- gridExtra::grid.arrange(p_3, p_2, p_1, nrow = 1)
```
