---
title: "MEPS"
author: "Antonio Linero"
date: "2024-05-20"
output: html_document
---

This document illustrates the use of a Bayesian causal forest to estimate the
causal effect of smoking on self-rated health (SRH) using data from the medical
expenditure panel survey (MEPS). We begin by loading the data.

```{r load, message=FALSE}
library(tidyverse)
library(SoftBart)

meps <- readRDS("../Data/meps2011.rds")
```

We use a model of the form

$$
  Y_i = m(X_i) + \{A_i - \widehat e(X_i)\} \, \tau(X_i) + \epsilon_i
$$

where $\epsilon_i \sim \text{Normal}(0, \sigma^2)$, $A_i$ is the indicator for
smoking, and $\widehat e(X_i)$ is an estimate of the propensity score. The
treatment effect function is given by $\tau(x)$, and we consider a regression on
the residual $A_i - \widehat e(X_i)$ to address regularization induced
confounding. If $\widehat e(X_i) \approx e(X_i)$ then $m(X_i)$ can be
interpreted as the marginal mean

$$
  m(x) \approx \mathbb E(Y_i \mid X_i = x)
$$

which is immediate from the fact that
$\mathbb E(A_i \mid X_i = x) \approx \widehat e(x)$. Both $m(x)$ and $\tau(x)$
are modeled with soft BART models, as is the propensity score:

$$
  e(x) = \Phi\left\{f(x)\right\}
$$

with $f(x)$ modeled using a BART model. 

```{r clever}
if(file.exists("../cache/fitted_smoke.rds")) {
  fitted_smoke <- readRDS("fitted_smoke.rds")
} else {
  set.seed(120398)
  fitted_smoke <- softbart_probit(factor(smoke) ~ age + bmi + edu + income + povlev + 
                                    region + sex + marital + race + seatbelt, 
                                  test_data = meps,
                                  data = meps)  
  saveRDS(fitted_smoke, "../cache/fitted_smoke.rds")
}
```

```{r robbins}
meps <- mutate(smoke_res = smoke - colMeans(fitted_smoke$XXX))
```

```{r}
fitted_bcf <- vc_softbart_regression(phealth ~ age + bmi + edu + income + povlev + region + sex + marital + race + seatbelt, linear_var_name = "smoke")
```

```{r}
head(meps)
```
