---
title: "MEPS"
author: "Antonio Linero"
date: "2024-05-20"
output: html_document
---

This document illustrates the use of a Bayesian causal forest to estimate the
causal effect of smoking on self-rated health (SRH) using data from the medical
expenditure panel survey (MEPS). We begin by loading the data.

```{r load, message=FALSE}
library(tidyverse)
library(SoftBart)
library(bcf)
library(ggdist)
library(patchwork)
library(latex2exp)
library(rpart)
library(rpart.plot)
theme_set(theme_bw())

meps <- readRDS("../Data/meps2011.rds")
head(meps)
```

SRH (`phealth`) is measured on a Likert scale with five levels, and has been
coded so that higher levels correspond to better health. In principle it is
probably better to model `phealth` as ordinal, but for the sake of this example
we will treat it as numeric.

# Model Description

We use a model of the form

$$
  Y_i = m(X_i) + \{A_i - \widehat e(X_i)\} \, \tau(X_i) + \epsilon_i
$$

where $\epsilon_i \sim \text{Normal}(0, \sigma^2)$, $A_i$ is the indicator for
smoking, and $\widehat e(X_i)$ is an estimate of the propensity score. The
treatment effect function is given by $\tau(x)$, and we consider a regression on
the residual $A_i - \widehat e(X_i)$ to address regularization induced
confounding. If $\widehat e(X_i) \approx e(X_i)$ then $m(X_i)$ can be
interpreted as the marginal mean

$$
  m(x) \approx \mathbb E(Y_i \mid X_i = x)
$$

which is immediate from the fact that
$\mathbb E(A_i \mid X_i = x) \approx \widehat e(x)$. Both $m(x)$ and $\tau(x)$
are modeled with soft BART models, as is the propensity score:

$$
  e(x) = \Phi\left\{f(x)\right\}
$$

with $f(x)$ modeled using a BART model.

# Some Causal Questions

We are interested in answering the following high-level questions:

1.  Are there any sub-populations for which the effect of smoking on health
    seems to be larger (smaller) than the overall effect? In particular, how
    does this relate to age, socio-economic status, and race?
2.  What is the overall (average) effect of smoking on self-rated health?

# Getting the Clever Covariate

Because the following code takes a long time to run, I have it set up to cache
the results of the computation so that it only needs to be run once.

```{r clever}
if(file.exists("../cache/fitted_smoke.rds")) {
  fitted_smoke <- readRDS("../cache/fitted_smoke.rds")
} else {
  set.seed(120398)
  opts <- Opts(update_s = FALSE, num_burn = 2500, num_thin = 5, num_save = 1000)
  fitted_smoke <- softbart_probit(factor(smoke) ~ age + bmi + edu + income + povlev + 
                                    region + sex + marital + race + seatbelt, 
                                  test_data = meps,
                                  data = meps,
                                  opts = opts)  
  saveRDS(fitted_smoke, "../cache/fitted_smoke.rds")
}
```

After fitting the model, we save the residual to the variable `smoke_res` in the
MEPS dataset:

```{r robbins}
meps <- mutate(meps, smoke_res = smoke - fitted_smoke$p_train_mean)
```

# Fitting the BCF

Next, we fit the Bayesian causal forest. We once again cache the result to avoid
repeating computations on re-runs. This time, however, we will use the `bcf`
package on CRAN.

```{r}
if(file.exists("../cache/fitted_bcf.rds")) {
  fitted_bcf <- readRDS("../cache/fitted_bcf.rds")
} else {
  set.seed(20948)
  fitted_bcf <- bcf(y = y, z = z, x_control = x_control, pihat = pihat, 
                  n_chains = 1, n_threads = 1, nburn = 1000, nsim = 1000, 
                  nthin = 1, update_interval = 10, no_output = TRUE,
                  save_tree_directory = NULL)
  saveRDS(fitted_bcf, "../cache/fitted_bcf.rds")
}
```

# The Overall Effect

The overall effect of smoking can be assessed by computing the *average
treatment effect*

$$
  \bar \tau = \mathbb E\{Y_i(1) - Y_i(0)\} = \int \tau(x) \, f_X(x) \ dx
$$

where $f_X(x)$ is the density of the covariates. To avoid specifying a
parametric family for $f_X(x)$ we instead take it to be (a) discrete and (b)
supported at each of the observed covariate values so that $f_X(x) = \omega_x$
and

$$
  \bar \tau = \sum_{i = 1}^N \omega_{X_i} \, \tau(X_i).
$$

We then model the $\omega_x$'s using the *Bayesian bootstrap* so that the
posterior distribution of the weights is given by

$$
  (\omega_{X_1}, \ldots, \omega_{X_N}) \sim \text{Dirichlet}(1, \ldots, 1).
$$

The following code gets the posterior distribution of $\bar \tau$ and visualizes
it:

```{r}
get_tau_bar <- function(fit) {
  num_iter <- nrow(fit$tau)
  num_obs <- ncol(fit$tau)
  omega <- MCMCpack::rdirichlet(n = num_iter, alpha = rep(1, num_obs))
  out_df <- data.frame(iter = 1:num_iter, tau_bar = rowSums(omega * fit$tau))
  return(out_df)
}
tau_bar_df <- get_tau_bar(fitted_bcf)

p_1 <- ggplot(tau_bar_df, aes(x = iter, y = tau_bar)) + geom_line() + xlab("Iteration") + ylab(TeX("\\bar{\\tau}")) + theme_bw()
p_2 <- ggplot(tau_bar_df, aes(x = tau_bar)) + stat_halfeye(.width = c(.90, .95, .99)) + ylab("Posterior Density") + xlab(TeX("\\bar{\\tau}"))

p_1 + p_2
```

We find that, unsurprisingly, smoking is associated with worse health outcomes:
on average, the causal effect of smoking is estimated to be a reduction in
health of 0.275 units.

# Subgroup Discovery

Next, we probe the posterior to determine if there are any subgroups for whom
the effect of smoking is relatively large. We first use a decision tree
to find compelling subgroups:

```{r}
post_sum <- rpart(colMeans(fitted_bcf$tau) ~ age + bmi + edu + income + povlev + 
                                    region + sex + marital + race + seatbelt,
                  data = meps)
rpart.plot(post_sum)
```

The most important predictor appears to be race, with education, age, and region
also appearing in the subgroups. We will restrict attention to differences among
races for this illustration.

```{r}
get_race_effects <- function(fit) {
  regions <- unique(meps$race)
  iters <- nrow(fit$tau)
  out <- list()
  for(race in regions) {
    idx <- which(meps$race == race)
    tau_hat_race <- rowMeans(fit$tau[,idx])
    tau_bar <- rowMeans(fit$tau)
    out[[race]] <- data.frame(iteration = iters, race = race, Delta = tau_hat_race - tau_bar)
  }
  return(do.call(rbind, out))
}

race_effects <- get_race_effects(fitted_bcf_2)
ggplot(race_effects, aes(x = Delta, y = race, fill = race)) + 
  stat_halfeye()
```

Overall, there seems to evidence that (i) white individuals have more negative
effects of smoking on self-rated health, while (ii) there is some evidence that
the difference in SRH among black individuals who smoke and who do not is
somewhat smaller than would be predicted by the average effect. There is some
indication that other races also have report smaller dropoffs than average,
but they overlap considerably with a $0$ difference from the average in their
posteriors.

# The Effect of Age

The following plot shows how we can also look at the treatment effects in
different age strata, as the effect of smoking might change depending on
an individual's age.

```{r}
age_brackets <- cut(meps$age, breaks = c(0, quantile(meps$age, seq(.1, 1, length = 20)), 120))
get_age_effects <- function(fit) {
  iters <- nrow(fit$tau)
  out <- list()  
  for(age in unique(age_brackets)) {
    idx <- which(age_brackets == age)
    tau_hat_age <- rowMeans(fit$tau[,idx])
    tau_bar <- rowMeans(fit$tau)
    out[[age]] <- data.frame(iteration = iters, 
                             age = age, 
                             Delta = tau_hat_age - tau_bar, 
                             tau_hat_age = tau_hat_age)
  }
  return(do.call(rbind, out))
}
age_effects <- get_age_effects(fitted_bcf)

ggplot(age_effects, aes(x = Delta, y = age, fill = age)) +
  stat_halfeye() + coord_flip()

```

Overahll, there seems to be some evidence that individuals are hit with larger
health impacts when they are older, but there is also substantial overlap
with $0$ at all ages.

