---
title: "An RCT"
author: "Antonio Linero"
date: "2024-05-20"
output: html_document
---

# Problem Setup

In this document we look at data from a clinical trial designed to assess
the efficacy of several different treatments for HIV/AIDS on survival time.
The study contained 2139 patients that were randomly assigned in equal
proportions to one of four treatment arms. To keep things simple, we will look
just at two of the treatments, which were coded as `trt == 1` and `trt == 2`.

First, we load the data:

```{r load}
library(SoftBart)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(ggdist)

aids <- read_csv("../Data/aids.csv")
aids_subset <- aids %>% filter(trt %in% c(0,1)) %>% 
  mutate(cd4_change = cd420 - cd40)
```

We take as our outcome the continuous measure of change in CD4 count from
baseline; CD4 count is a commonly-used biomarker that tracks the progression
of AIDS symptoms, and generally individuals with lower CD4 counts have a poorer
prognosis. Below we give a boxplot of the change from baseline CD4 across the
two treatments.

```{r}
ggplot(aids_subset, aes(x = factor(trt), y = cd4_change)) + geom_boxplot() + theme_bw()
```

We can see from this plot a fairly noticeable shift in CD4 counts between the
treatment and control arm, with CD4 changes tending to be somewhat higher in
the treatment arm.

Because the trial was randomized, we do not have to worry about confounding
between the treatment assignment and the pretreatment variables. We will
threfore just use the `vc_softbart_regression` function with a linear predictor
equal to $A_i - 0.5$.

```{r}
set.seed(992093)

formula_aids <- cd4_change ~ symptom + age + wtkg + hemo + homo + drugs + 
  karnof + oprior + z30 + zprior + preanti + race + gender + str2 + symptom

my_fit <- vc_softbart_regression(
  formula_aids,
  linear_var_name = "trt",
  data = aids_subset %>% mutate(trt = trt - 0.5),
  test_data = aids_subset %>% mutate(trt = trt - 0.5),
  num_tree = 50,
  k = 1,
  opts = Opts(
    num_burn = 1000,
    num_save = 1000,
    num_thin = 5,
    update_s = FALSE
  )
)
```

# The Overall Treatment Effect

First we will use the model to estimate the overall treatment effect. This is
mainly just a sanity check: because the trial is randomized, we could also just
do a t-test for the difference. Below we give the average treatment effect:

```{r}
plotdf <- data.frame(iteration = 1:nrow(my_fit$beta_train), 
                     ATE = rowMeans(my_fit$beta_train))
ggplot(data = plotdf, aes(x = ATE)) + stat_halfeye() + theme_bw()
plotdf %>% summarise(ate_hat = mean(ATE), 
                     LCL = quantile(ATE, 0.025), 
                     UCL = quantile(ATE, 0.975))
```

This is comperable with what we get from the t-test:

```{r}
t.test(cd4_change ~ trt, data = aids_subset) %>% tidy()
```

# Treatment Effect Heterogeneity

First, we investigate treatment effect heterogeneity using subgroup
identification, more-or-less just to check whether there is much here worth
examining. First we run virtual twins using the `rpart` and `rpart.plot`
packages.

```{r}
tau_samps <- my_fit$beta_train
tau_hats <- colMeans(tau_samps)
subgroup_rpart <- rpart(tau_hats ~ symptom + age + wtkg + hemo + homo + drugs + 
  karnof + oprior + z30 + zprior + preanti + race + gender + str2 + symptom,
  data = aids_subset)
rpart.plot(subgroup_rpart)
```

We see that the tree produces remarkably aligned subgroups, focusing primarily
on young-vs-old at the first split, sexual orientation at the second split,
and race at the third split. Let's focus just on the first two splits:

```{r}
g_11 <- with(aids_subset, which(age < 37 & homo == 1))
g_10 <- with(aids_subset, which(age < 37 & homo == 0))
g_01 <- with(aids_subset, which(age >= 37 & homo == 1))
g_00 <- with(aids_subset, which(age >= 37 & homo == 0))

tau_hat_11 <- rowMeans(my_fit$beta_train[,g_11])
tau_hat_10 <- rowMeans(my_fit$beta_train[,g_10])
tau_hat_01 <- rowMeans(my_fit$beta_train[,g_01])
tau_hat_00 <- rowMeans(my_fit$beta_train[,g_00])
tau_hat_all <- rowMeans(my_fit$beta_train)
iteration <- 1:length(tau_hat_11)
delta_hat_11 <- tau_hat_11 - tau_hat_all
delta_hat_10 <- tau_hat_10 - tau_hat_all
delta_hat_01 <- tau_hat_01 - tau_hat_all
delta_hat_00 <- tau_hat_00 - tau_hat_all

plot_df <-
  data.frame(
    tau_hat = c(tau_hat_11, tau_hat_10, tau_hat_01, tau_hat_00),
    delta_hat = c(delta_hat_11, delta_hat_10, delta_hat_01, delta_hat_00),
    iteration = rep(iteration, 4),
    subgroup = rep(c("g11", "g10", "g01", "g00"), each = length(iteration))
  )

ggplot(plot_df, aes(x = tau_hat, y = subgroup)) + stat_halfeye()
ggplot(plot_df, aes(x = delta_hat, y = subgroup)) + stat_halfeye() + 
  theme_bw() + geom_vline(xintercept = 0)
```

On the basis of this, we see that there is some marginal evidence that the
treatment was, on average, less effective within the group of young homosexual
individuals and more effective within the group of older heterosexual
individuals. Further refining by race gives the following:

```{r}
g_111 <- with(aids_subset, which(age < 37 & homo == 1 & race == 1))
g_101 <- with(aids_subset, which(age < 37 & homo == 0 & race == 1))
g_011 <- with(aids_subset, which(age >= 37 & homo == 1 & race == 1))
g_001 <- with(aids_subset, which(age >= 37 & homo == 0 & race == 1))
g_110 <- with(aids_subset, which(age < 37 & homo == 1 & race == 0))
g_100 <- with(aids_subset, which(age < 37 & homo == 0 & race == 0))
g_010 <- with(aids_subset, which(age >= 37 & homo == 1 & race == 0))
g_000 <- with(aids_subset, which(age >= 37 & homo == 0 & race == 0))

tau_hat_111 <- rowMeans(my_fit$beta_train[,g_111])
tau_hat_101 <- rowMeans(my_fit$beta_train[,g_101])
tau_hat_011 <- rowMeans(my_fit$beta_train[,g_011])
tau_hat_001 <- rowMeans(my_fit$beta_train[,g_001])
tau_hat_110 <- rowMeans(my_fit$beta_train[,g_110])
tau_hat_100 <- rowMeans(my_fit$beta_train[,g_100])
tau_hat_010 <- rowMeans(my_fit$beta_train[,g_010])
tau_hat_000 <- rowMeans(my_fit$beta_train[,g_000])

tau_hat_all <- rowMeans(my_fit$beta_train)
iteration <- 1:length(tau_hat_11)
delta_hat_111 <- tau_hat_111 - tau_hat_all
delta_hat_101 <- tau_hat_101 - tau_hat_all
delta_hat_011 <- tau_hat_011 - tau_hat_all
delta_hat_001 <- tau_hat_001 - tau_hat_all
delta_hat_110 <- tau_hat_110 - tau_hat_all
delta_hat_100 <- tau_hat_100 - tau_hat_all
delta_hat_010 <- tau_hat_010 - tau_hat_all
delta_hat_000 <- tau_hat_000 - tau_hat_all

plot_df <-
  data.frame(
    tau_hat = c(tau_hat_111, tau_hat_101, tau_hat_011, tau_hat_001, 
                tau_hat_110, tau_hat_100, tau_hat_010, tau_hat_000),
    delta_hat = c(delta_hat_111, delta_hat_101, delta_hat_011, delta_hat_001, 
                  delta_hat_110, delta_hat_100, delta_hat_010, delta_hat_000),
    iteration = rep(iteration, 8),
    subgroup = rep(c("g111", "g101", "g011", "g001", 
                     "g110", "g100", "g010", "g000"), each = length(iteration))
  )

ggplot(plot_df, aes(x = tau_hat, y = subgroup)) + stat_halfeye()
ggplot(plot_df, aes(x = delta_hat, y = subgroup)) + stat_halfeye() + 
  theme_bw() + geom_vline(xintercept = 0)
```

Of course, how far down the tree we go should depend on (i) the sample size (as
the subgroups get smaller, the intervals will be wider), (ii) the magnitude of
the effects (larger effects will make the effects more statistically plausible),
and (iii) how much we believe in the regularization provided by the prior (if we
do not believe that the prior is properly calibrated, and are just hoping for
Frequentsit valid inferences, then we are effectively just data dredging as we
see to find "statistically significant" effects).

Overall, I would conclude on the basis of this work that the some examination of
the treatment effect across age/race/sexual orientation warrants further study,
in particular in trying to identify a mechanism for any apparent effects, but
the analysis does not provide firm evidence of large differences. The effects
differences *are* somewhat large: for the coarsest subgroups we get differences
on the scale of 25% of the main effect.

For reference, this is the result of an ANOVA that looks specifically at these
three factors:

```{r}
aids_subset$age_cut <- ifelse(aids_subset$age > 37, 1, 0)
anova_lm <- lm(cd4_change ~ trt * (age_cut + race + homo), data = aids_subset)
anova_add <- lm(cd4_change ~ age_cut + race + homo + trt, data = aids_subset)

anova(anova_add, anova_lm)
```

The results of this analysis suggests a statistically significant interaction
between the treatment and some combination of these three variables; the catch
of course here as far as Frequentist-validity of these results is that we have
effectively "used the data twice" in arriving at this conclusion, whereas in
some sense the BART model uses the data "only once" in the move from the prior
to the posterior.
