---
title: "An RCT"
author: "Antonio Linero"
date: "2024-05-20"
output: html_document
---

In this document we look at data from a clinical trial designed to assess
the efficacy of several different treatments for HIV/AIDS on survival time.
The study contained 2139 patients that were randomly assigned in equal
proportions to one of four treatment arms. To keep things simple, we will look
just at two of the treatments, which were coded as `trt == 1` and `trt == 2`.

First, we load the data:

```{r load}
library(SoftBart)
library(tidyverse)
library(rpart)
library(rpart.plot)

aids <- read_csv("../Data/aids.csv")
aids_subset <- aids %>% filter(trt %in% c(0,1)) %>% 
  mutate(cd4_change = cd420 - cd40)
```

Because the trial was randomized, we do not have to worry about confounding
between the treatment assignment and the XXX

```{r}
set.seed(992093)

formula_aids <- cd4_change ~ symptom + age + wtkg + hemo + homo + drugs + 
  karnof + oprior + z30 + zprior + preanti + race + gender + str2 + symptom

my_fit <- vc_softbart_regression(
  formula_aids,
  linear_var_name = "trt",
  data = aids_subset %>% mutate(trt = trt - 0.5),
  test_data = aids_subset %>% mutate(trt = trt - 0.5),
  num_tree = 50,
  k = 1,
  opts = Opts(
    num_burn = 1000,
    num_save = 1000,
    num_thin = 5,
    update_s = FALSE
  )
)
```

```{r}
plot()
```

# Trying with a BCF

```{r}
library(bcf) 

set.seed(23832)

fitbcf <- function() {
  X <- model.matrix(~ symptom + age + wtkg + hemo + homo + drugs +  karnof + 
                      oprior + z30 + zprior + preanti + race + gender + str2 + 
                      symptom, 
                    data = aids_subset)
  trt <- aids_subset$trt
  y <- aids_subset$cd4_change
  
  bcf_out <- bcf(y = y, z = trt, x_control = X, x_moderate = X, 
                 pihat = rep(0.5, nrow(X)), nburn = 1000, nsim = 1000, 
                 no_output = TRUE, log_file = NULL, n_chains  = 1)
  return(bcf_out)
}

fitted_bcf <- fitbcf()
```

## Plotting

```{r}
plot(colMeans(fitted_bcf$yhat), aids_subset$cd4_change)
abline(a=0,b=1)
hist(rowMeans(fitted_bcf$tau))
```

## Output of BCF

```{r}
race_idx <- which(aids_subset$homo == 1 & aids_subset$race == 1 & aids_subset$drugs == 0)
race_idx2 <- which(aids_subset$homo == 0 & aids_subset$race == 0)

tau_race_1 <- rowMeans(fitted_bcf$tau[,race_idx])
tau_race_0 <- rowMeans(fitted_bcf$tau[,race_idx2])

hist(tau_race_1 - tau_race_0)
```

```{r}
# tau_hat <- colMeans(my_fit$beta_train)
tau_hat <- colMeans(fitted_bcf$tau)
rpart.plot(rpart(tau_hat ~ symptom + age + wtkg + hemo + homo + drugs + 
  karnof + oprior + z30 + zprior + preanti + race + gender + str2 + symptom, 
  data = aids_subset))
```

## What if We Just Use a Linear Model?

```{r}
bclm <- lm(cd4_change ~ trt * race * drugs, data = aids_subset)
summary(bclm)
step(bclm) %>% summary()
```

## What if we try CoxPH?

Seems like not much evidence that there is anything interesting going on in
terms of survival rates for the different results; treatment seems to be
highly effective. What about the whole dataset?

```{r}
library(survival)
coxfit <- coxph(Surv(time, label) ~ trt * (race + drugs + homo + gender), data = aids_subset)
summary(coxfit)

coxfit2 <- coxph(Surv(time, label) ~ factor(trt) * (race + drugs + homo + gender), data = aids)
summary(coxfit2)
step(coxfit2)
```

Still doesn't really seem to support anything interesting going on vis-a-vis 
demographics or whatnot.